---
title: "Exponential Distributions"
author: "Alex Petkovski"
date: "Saturday, May 23, 2015"
output:
  pdf_document:
    latex_engine: xelatex
---
# Overview   
The following report investigates properties of the **Exponential Distribution** 
in **R** wherein we will compare it with the ***Central Limit Theorem*** and we will 
set **lambda = 0.2** for all simulations. Using means of 40 exponentials from a
sample of 1000, we will compare **sample vs. theoretical mean** and **sample vs. 
theoretical variance**. Finally, we will compare the difference between a large 
collection of **random exponentials** and the distribution of a large collection 
of **averages of 40 exponentials** and show that it is approximately normal.

# Simulations   

## Sample Means   
- Simulate $n = 40$ exponentials a 1000 times
- Let $X_i$ be the outcome for sample $i$
- Then note that $\mu = E[X_i] = 1/\lambda = 1/0.2 = 5$ (Vertical dotted line)
- Let's take $n$ exponentials, take their mean and repeat this 1000 times 
(Simulation Means)   
```{r, fig.height=4, figh.width=6, fig.align='center', echo = FALSE, cache = TRUE}
library(ggplot2)
nosim <- 1000; n <- 40; lambda = 0.2;
tmean <- 1/(lambda);
dmean <- data.frame(
    x = c(apply(matrix(rexp(nosim * n, lambda), nosim), 1, mean))
    )
ggplot(dmean, aes(x = x, fill = "Simulation Means")) +
    geom_density(size = 2, alpha = .2) + 
    geom_vline(xintercept = tmean, size = 2, linetype = 3) +
    theme(legend.position="bottom", legend.title=element_blank()); 
smean <- mean(dmean$x)
```
   
## Sample Variances   
- Simulate $n = 40$ exponentials a 1000 times
- Let $X_i$ be the outcome for sample $i$
- Then note that $\mu = E[X_i] = 1/\lambda^2 = 1/0.2^2 = 25$ (Vertical dotted line)
- Let's take $n$ exponentials, take their mean and repeat this 1000 times 
(Simulation Variances)
```{r, fig.height=4, figh.width=6, fig.align='center', echo = FALSE, cache = TRUE}
library(ggplot2)
nosim <- 1000; n <- 40; lambda = 0.2;
tvar <- 1/(lambda^2);
dvar <- data.frame(
    x = c(apply(matrix(rexp(nosim * n, lambda), nosim), 1, var))
    )
ggplot(dvar, aes(x = x, fill = "Simulation Variances")) +
    geom_density(size = 2, alpha = .2) + 
    geom_vline(xintercept = tvar, size = 2, linetype = 3) +
    theme(legend.position="bottom", legend.title=element_blank()); 
svar <- mean(dvar$x)
```

## Normal Approximation of the Distribution
- Simulate $n = 40$ exponentials a 1000 times
- Let $X_i$ be the outcome for sample $i$
- Then note that $\mu = E[X_i] = 1/\lambda = 1/0.2 = 5$
- $Var(X_i) = 1 / \lambda^2 = 25$ 
- SE $\sqrt{25 / n} = 5 / \sqrt{n}$
- Let's take $n$ exponentials, take their mean, subtract off 5,
and divide by $5 / \sqrt{n}$ and repeat this 1000 times (Simulation Histogram)
- Compare this against normal distribution (Dotted curve)
```{r, echo = FALSE, fig.width=6, fig.height = 4, fig.align='center', cache=TRUE}
library(ggplot2)
nosim <- 1000; n <- 40; lambda = 0.2;
cfunc <- function(x, n) sqrt(n) * (mean(x) - 1/lambda) / sqrt(1/lambda^2)
dat <- data.frame(
  x = apply(matrix(rexp(nosim * n, lambda), 
                     nosim), 1, cfunc, n)
  )
g <- ggplot(dat, aes(x = x)) + 
    geom_histogram(alpha = .20, 
                   binwidth=.3, 
                   colour = "black", 
                   fill = "pink",
                   aes(y = ..density..)) 
g <- g + stat_function(fun = dnorm, size = 2, linetype = 3)
g
s2mean <- mean(dat$x)
s2var <- var(dat$x)
```

# Sample Mean vs Theoretical Mean  
- Mean of simulation means is `r smean`
- Theoretical mean is `r tmean`
- Thus simulation means are a good estimate of the mean and from the first plot 
we can see that it is an unbiased estimate the mean.

# Sample Variance vs Theoretical Variance  
- Mean of simulation variances is `r svar`
- Theoretical variance is `tvar`
- Thus simulation variances are a good estimate of the variance and from the 
second plot we can see that it is an unbiased estimate of the variance.

# Distribution
- Mean of normalized simulation is `r s2mean`
- Variance of normalized simulation is `r s2var`
- These are close to `0` and `1` respectively which shows that the **Central Limit 
Theorem** (CLT) is correct for the exponential distribution, i.e. the distribution 
of averages of iid variables becomes that of the standard normal as the sample 
size increases.

# Appendix
The following shows R code used in the analysis:   

- Code that generates the first graph and calculates the simulation and theoretical 
mean.   
```
library(ggplot2)
nosim <- 1000; n <- 40; lambda = 0.2;
tmean <- 1/(lambda);
dmean <- data.frame(
    x = c(apply(matrix(rexp(nosim * n, lambda), nosim), 1, mean))
    )
ggplot(dmean, aes(x = x, fill = "Simulation Means")) +
    geom_density(size = 2, alpha = .2) + 
    geom_vline(xintercept = tmean, size = 2, linetype = 3) +
    theme(legend.position="bottom", legend.title=element_blank()); 
smean <- mean(dmean$x)
```

- Code that generates the second graph and calculates the simulation and theoretical
variance.   
```
library(ggplot2)
nosim <- 1000; n <- 40; lambda = 0.2;
tvar <- 1/(lambda^2);
dvar <- data.frame(
    x = c(apply(matrix(rexp(nosim * n, lambda), nosim), 1, var))
    )
ggplot(dvar, aes(x = x, fill = "Simulation Variances")) +
    geom_density(size = 2, alpha = .2) + 
    geom_vline(xintercept = tvar, size = 2, linetype = 3) +
    theme(legend.position="bottom", legend.title=element_blank()); 
svar <- mean(dvar$x)
```

- Code that generates the third graph and calculates the normalized simulation mean
and variance.   
```
library(ggplot2)
nosim <- 1000; n <- 40; lambda = 0.2;
cfunc <- function(x, n) sqrt(n) * (mean(x) - 1/lambda) / sqrt(1/lambda^2)
dat <- data.frame(
  x = apply(matrix(rexp(nosim * n, lambda), 
                     nosim), 1, cfunc, n)
  )
g <- ggplot(dat, aes(x = x)) + 
    geom_histogram(alpha = .20, 
                   binwidth=.3, 
                   colour = "black", 
                   fill = "pink",
                   aes(y = ..density..)) 
g <- g + stat_function(fun = dnorm, size = 2, linetype = 3)
g
s2mean <- mean(dat$x)
s2var <- var(dat$x)
```
