---
title: "Tooth Growth"
author: "Alex Petkovski"
date: "Sunday, May 24, 2015"
output:
  pdf_document:
    latex_engine: xelatex
---

# Overview
The following report investigates **R** dataset **Tooth Growth**. Some preliminary 
exploratory data analysis is conducted along with a summary of the key statistics. 
Finally, we construct some hypothesis tests to compare tooth growth by supplement 
factor and dose in order to state some conclusions and assumptions needed for those 
conclusions.

# Data Exploration
Load the data, get the structure and summarize
```{r, cache=TRUE}
data(ToothGrowth)
```
```{r, echo=FALSE, cache=TRUE}
str(ToothGrowth)
summary(ToothGrowth)
```
   
From the structure and review of `?ToothGrowth` help file, we have the following
structure and description:   

Variable | Data Type | Description
-------- | --------- | --------------------------
len      | numeric   | Tooth length
supp     | factor    | Supplement type (VC or OJ)
dose     | numeric   | Dose in milligrams

Let's make **dose** a factor since we will be analyzing tooth length by dose 
factor
```{r, cache=TRUE}
ToothGrowth$dose <- as.factor(ToothGrowth$dose)
```

Finally, let's review the data in a Supplement type panel box plot   
```{r, echo=FALSE, cache=TRUE}
library(ggplot2)
ggplot(ToothGrowth, aes(x=dose, y=len, fill=dose)) + 
    geom_boxplot() + 
    facet_grid(. ~ supp) +
    guides(fill=FALSE)
```

# Summary
Assuming a normal distribution, we'll construct a table that calculates the various 
statistics such as $X_bar$, $S$ and $n$.  
```{r, echo=FALSE, cache=TRUE}
attach(ToothGrowth)
dsum <- aggregate(len,list(dose,supp),FUN=function(x) c(X_bar = mean(x)
                                                       , S = sd(x)
                                                       , n = length(x)))
names(dsum) <- c("dose","supp","x")
dsum
detach(ToothGrowth)
```

# Confidence Intervals and Hypothesis Tests

## Supplement Type Investigation
Does supplement type (VC or OJ) have any impact on tooth length?
- H0: different supplement types have no effect on tooth length
```{r, echo=FALSE, cache=TRUE}
t.test(len ~ supp, paired = TRUE, data = ToothGrowth)
```

t is high, confidence does not contain 0 and p-value is less that 5%, thus we 
reject H0.

## Dose Investigation
Does dose have any impact on tooth length?
- H0: different doses have no effect on tooth length
```{r, echo=FALSE, cache=TRUE}
d_05 <- ToothGrowth$len[ToothGrowth$dose == 0.5]
d_10 <- ToothGrowth$len[ToothGrowth$dose == 1]
d_20 <- ToothGrowth$len[ToothGrowth$dose == 2]

t_05v10 <- t.test(d_05, d_10, paired = TRUE)
t_05v20 <- t.test(d_05, d_20, paired = TRUE)
t_10v20 <- t.test(d_10, d_20, paired = TRUE)

t1 <- data.frame(
    "t.value" = sprintf("%.3f", t_05v10$statistic),  
    "CI-low" = sprintf("%.3f", t_05v10$conf[1]),
    "CI-high" = sprintf("%.3f", t_05v10$conf[2]),
    "p.value" = sprintf("%.3f", t_05v10$p.value)
)
t2 <- data.frame(
    "t.value" = sprintf("%.3f", t_05v20$statistic),  
    "CI-low" = sprintf("%.3f", t_05v20$conf[1]),
    "CI-high" = sprintf("%.3f", t_05v20$conf[2]),
    "p.value" = sprintf("%.3f", t_05v20$p.value)
)
t3 <- data.frame(
    "t.value" = sprintf("%.3f", t_10v20$statistic),  
    "CI-low" = sprintf("%.3f", t_10v20$conf[1]),
    "CI-high" = sprintf("%.3f", t_10v20$conf[2]),
    "p.value" = sprintf("%.3f", t_10v20$p.value)
)

result <- rbind(t1, t2, t3)
result

```

t is very low, confidence does not contain 0 and p-value is less that 5%, thus we 
reject H0 across all tests.

# Conclusions
Assuming that t-tests are paired which is based on description of Tooth Growth
data-set, we conclude the following:
- Reject hypothesis that supplement types have no effect on tooth growth.
- Reject hypothesis that doses have no effect on tooth growth.
- Both conclusions also assume that two variances in all t-tests are not equal.

# Appendix

Code used to get structure and summary of data
```
str(ToothGrowth)
summary(ToothGrowth)
```

Code used to generate box-plot
```
library(ggplot2)
ggplot(ToothGrowth, aes(x=dose, y=len, fill=dose)) + 
    geom_boxplot() + 
    facet_grid(. ~ supp) +
    guides(fill=FALSE)
```

Code used to generate summary table
```
attach(ToothGrowth)
dsum <- aggregate(len,list(dose,supp),FUN=function(x) c(X_bar = mean(x)
                                                       , S = sd(x)
                                                       , n = length(x)))
names(dsum) <- c("dose","supp","x")
dsum
detach(ToothGrowth)
```

Code for supplement test
```
t.test(len ~ supp, paired = TRUE, data = ToothGrowth)
```

Code for dose test
```
d_05 <- ToothGrowth$len[ToothGrowth$dose == 0.5]
d_10 <- ToothGrowth$len[ToothGrowth$dose == 1]
d_20 <- ToothGrowth$len[ToothGrowth$dose == 2]

t_05v10 <- t.test(d_05, d_10, paired = TRUE)
t_05v20 <- t.test(d_05, d_20, paired = TRUE)
t_10v20 <- t.test(d_10, d_20, paired = TRUE)

t1 <- data.frame(
    "t.value" = sprintf("%.3f", t_05v10$statistic),  
    "CI-low" = sprintf("%.3f", t_05v10$conf[1]),
    "CI-high" = sprintf("%.3f", t_05v10$conf[2]),
    "p.value" = sprintf("%.3f", t_05v10$p.value)
)
t2 <- data.frame(
    "t.value" = sprintf("%.3f", t_05v20$statistic),  
    "CI-low" = sprintf("%.3f", t_05v20$conf[1]),
    "CI-high" = sprintf("%.3f", t_05v20$conf[2]),
    "p.value" = sprintf("%.3f", t_05v20$p.value)
)
t3 <- data.frame(
    "t.value" = sprintf("%.3f", t_10v20$statistic),  
    "CI-low" = sprintf("%.3f", t_10v20$conf[1]),
    "CI-high" = sprintf("%.3f", t_10v20$conf[2]),
    "p.value" = sprintf("%.3f", t_10v20$p.value)
)

result <- rbind(t1, t2, t3)
result

```
